---
title: "MH3511 Lab 1 Notes"
author: "Caleb Tay (caleb.tay@ntu.edu.sg)"
date: "22 Jan 2026"
output:
  html_document:
    toc: true
    toc_float: true
  pdf_document:
    df_print: tibble
    template: ../eisvogel.latex
toc: true
toc-own-page: true
numbersections: true
papersize: a4
geometry:
- top=25mm
- left=25mm
- right=25mm
- bottom=25mm
- headsep=4mm
- footskip=10mm
header-includes:
   - \usepackage{graphicx}
editor_options:
  markdown:
    wrap: 80
---

# Exercise 1.1

```{r results='hold'}
205 + 106
2 * 6 + 9 / 3
2 * (6 + 9) / 3
3^4 - 4^3
28 %% 6
```

There is a general order of operations, which follows regular math.

# Exercises 1.2

```{r}
red_apples <- 4
green_apples <- 5
basket_apples <- red_apples + green_apples
18 * basket_apples
```

While you could do (c) with `basket_apples <- 4 + 5`, this makes the code less
flexible when the variables `red_apples` and `green_apples` are representing a
specific quantity, i.e. the number of red apples and green apples in the basket.

For example, we can easily change the values of `red_apples` and `green_apples`
and get the total sum, changing the values at *only one place*.

# Exercise 1.3

```{r}
my_course <- "MH1234"
my_score <- 76
my_grade <- "A-"
my_core <- TRUE

class(my_course)
class(my_score)
class(my_grade)
class(my_core)
```

Boolean values can be written with a single letter, i.e. `T` for `TRUE` and `F` for `FALSE`.

# Exercise 1.4

```{r}
seq(10)
seq(0, 10, length = 10)
seq(0, 10, by = 1)
rep(2, 3)
rep(-1, 4)
rep(1:3, 2)
rep("A", 4)
c(rep("A", 3), rep("B", 3))
rep(c("A", "B"), 3)
```

```{r}
vector1 <- seq(3, 9, 2)
vector2 <- rep(2:4, 2:4)
vector3 <- c(rep(T, 2), rep(F, 3))
vector3 <- rep(c(T, F), c(2, 3))
vector4 <- rep(c("High", "Medium", "Low"), 2)
vector1; vector2; vector3; vector4
```

# Exercise 1.5

```{r}
poker_winnings <- c(140, -50, 20, -120, 240)
roulette_winnings <- c(-24, -50, 100, -350, 10)
days <- c("Mon", "Tue", "Wed", "Thur", "Fri")

poker_winnings
names(poker_winnings)
names(poker_winnings) <- days
poker_winnings

names(roulette_winnings) <- days
poker_winnings
poker_winnings > 0
poker_winnings[poker_winnings > 0]

daily_winnings <- poker_winnings + roulette_winnings
daily_winnings
daily_winnings > 0
daily_winnings[daily_winnings > 0]

sum(poker_winnings)
sum(roulette_winnings)
```

(j): Based on the summations, we can clearly see poker had a net gain, while
roulette had a net loss. Clearly, don't play roulette.

```{r}
sum(daily_winnings)
```

(k): There was a net loss of \$84 in the week.

# Exercise 1.6

```{r}
matrix_4x2 <- matrix(1:8, ncol = 2)
matrix_4x2

matrix_2x4 <- matrix(c(1:4, 11:14), nrow = 2, byrow=T)
matrix_2x4

matrix_4x2 %*% matrix_2x4
matrix_2x4 %*% matrix_4x2
```

```{r}
matrix_4x2
matrix_2x2 <- matrix_4x2[1:2,]
matrix_2x2
```

Note that matrix multiplication is done using `%*%` in R. The `solve()` function
attempts to find the inverse of a square matrix. To verify it is the inverse, we
can perform matrix multiplication.

```{r}
solve(matrix_2x2)
matrix_2x2 %*% solve(matrix_2x2)

diag(3)
diag(c(2, 3, 5))
diag(matrix_2x2)
```

# Exercise 1.7

```{r}
apple <- c(350, 150)
orange <- c(400, 230)
watermelon <- c(180, 90)

fruit_matrix <- matrix(c(apple, orange, watermelon), nrow = 3, byrow = T)

fruit_matrix
rownames(fruit_matrix) <- c("Apple", "Orange", "Watermelon")
colnames(fruit_matrix) <- c("Weekday", "Weekend")
fruit_matrix

sum_fruit_matrix <- cbind(fruit_matrix, Total=rowSums(fruit_matrix))
sum_fruit_matrix
```

Above, I add the weekday and weekend totals into a new matrix just so it is
easier to run repeatedly.

```{r}
sum_fruit_matrix[,"Total"] * c(1, 1.5, 4)
sum(sum_fruit_matrix[,"Total"] * c(1, 1.5, 4))

sum_fruit_matrix[,"Total"] %*% c(1, 1.5, 4)
c(1, 1.5, 4) %*% sum_fruit_matrix[,"Total"]
```

The `%*%` operator here is acting on two *vectors*, so the inner product of the
two vectors is obtained instead.

# Getting help on matrix multiplication

As a side note, to get help on matrix multiplication (or any other operator),
you have to surround it with backticks (\`), typically found above the tab key,
with the tilde (\~).

```{r eval=FALSE}
?`%*%`
help(`%*%`)
```
